Accuracy in Parameter Estimation and Simulation Approaches for Sample Size Planning Accounting for Item Effects

### 2024-12-01 manuscript_draft.Rmd
updating manuscript draft

In this section, we provide an example of the suggested procedure. The first dataset includes concreteness ratings from @brysbaert2014. Instructions given to participants denoted the difference between concrete (i.e., "refers to something that exists in reality") and abstract (i.e., "something you cannot experience directly through your senses or actions") terms. Participants were then asked to rate concreteness of terms using a 1 (*abstract*) to 5 (*concrete*) scale. This data represents a small scale dataset (i.e., the range of the scale of the data is small, 4 points) that could be used as pilot data for a study using concrete word ratings. The data is available at <https://osf.io/qpmf4/>.

@@

For this example, imagine a researcher who wants to determine the differences in response latencies for abstract and concrete words [see the following for thoughts on analyzing this type of data: @bürkner2019; @liddell2018; @taylor2022]. They will select *n* = 40 words from the rating data from @brysbaert2014 that are split evenly into abstract and concrete ends of the rating scale. In the experiment, each participant will be asked to rate the words for their concreteness, and then complete a lexical decision task with these words as the phenomenon of interest. Using both datasets and the procedure outlined above, we can determine the sample size necessary to ensure adequately measured concreteness ratings and response latencies.

@@

*Step 1*. The concreteness ratings data includes `r length(unique(conDF$Word))` concepts that were rated for their concreteness. We randomly selected *n* = 20 abstract words ($M_{Rating}$ \<= 2) and *n* = 20 concrete words ($M_{Rating}$ \>= 4). In the original study, not every participant rated every word, which created uneven sample sizes for each word. Further, participants were allowed to indicate they did not know a word, and those responses were set to missing data. In our sample of 40 words, the average pilot sample size was `r round(mean(tapply(con_small$Rating, con_small$Word, length)), digits = 2)` (*SD* = `r round(sd(tapply(con_small$Rating, con_small$Word, length)), digits = 2)`), and we will use `r pilot_size_c` as our pilot sample size for the concreteness ratings (this information will be used in the follow-up to the simulation study).

@@

We first selected the same real words in the ELP data as the concreteness subset selected above, and this data includes `r length(unique(elpDF$Stimulus))` real words. The average pilot sample size for this random sample was `r format(round(mean(tapply(elp_small$RT, elp_small$Stimulus, length)), digits = 2), nsmall = 2)` (*SD* = `r round(sd(tapply(elp_small$RT, elp_small$Stimulus, length)), digits = 2)`), and *n* = `r pilot_size_e` will be our pilot size for the lexical decision task.

@@

## Special Consideration: Bimodal Data

The above simulations assume symmetric population distributions with small, medium, or large heterogeneity or a skewed distribution with the same heterogeneity. However, as shown in @pollock2018, sample rating data that appears to have a mean distribution may represent the average two separate distributions (i.e., a bimodal distribution). To examine our procedure with bimodal distributions, we estimated thirty items for three distributions: symmetric, ceiling, and floor distributions for Likert data with small heterogeneity. Bimodal data was created by selecting half of the ceiling distribution and floor distribution to combine together. The number of bimodal items was varied from 0 percent to 100 percent increasing by 10 percent increments for simulated samples. Figure \@\ref(fig:bimodal-figure) shows an example of 10 of the items within a simulation that estimated that half of the Likert items would be bimodal in nature. The researcher procedure described above was then carried out using *semanticprimeR*.

```{r import-bimodal, echo = F}
bi_DF <- import("simulation_study/simulation_bimodal/summary_bimodal.csv") %>% 
  mutate(new_percent = rep(c(80, 85, 90, 95), times = 11)) %>% 
  mutate(size = factor(size*100))

@@

```{r bimodal-figure, fig.cap = "Simulated example of Likert data with half of the items as bimodal distributions.", fig.height=8, fig.width=8}
ggplot(researcher_sample %>% 
         filter(item < 11), aes(x = score, y = factor(item), fill = factor(item))) +
  geom_density_ridges(alpha = 0.7, scale = 1) +
  labs(x = "Score", y = "Item") +
  theme_minimal() +
  theme(legend.position = "none") + 
  coord_cartesian(xlim = c(1,7))
```

Figure \@\ref(fig:bimodal-figure) portrays the results of the suggested procedure. The sample with 0 bimodal items demonstrates the same values as above. The proposed sample size increases from 10 to 40% because of the heterogeneity in SE across items (i.e., with 10% of items with a larger SE due to their bimodal nature, the required sample size increases). From 50% to 100% the sample size decreases because the variance of the SEs across items decreases (i.e., at 100 percent, they are all larger rather than a mix). In theory, these results map onto the conceptual framework - if all items are truly bimodal, we have precisely measured the mean of the bimodal distribution, but this result is likely not the intended result. If researchers expect these distributions (or have representative pilot data), they could examine the data for bimodal items. These items could be estimated separately (i.e., floor effects item 1, ceiling effects item 1) to ensure that both populations of answers are represented in the data.

```{r bimodal-results, fig.cap = "Estimated sample sizes for distributions that vary in the number of bimodal items. Note that percent below is treated as a continuous factor", fig.height=8, fig.width=8}
ggplot(bi_DF, aes(x = percent_below,
                  y = corrected_sample_size, 
                  color = factor(size),
                  shape = factor(size))) +
  geom_point(size = 3) +
  geom_line() +
  theme_classic() +
  xlab("Percent of Items Below Threshold") + 
  ylab("Corrected Sample Size") +
  scale_color_viridis_d(option = "plasma") +  # Use the 'plasma' viridis option
  scale_shape_manual(values = c(16, 17, 18, 19, 15, 3, 7, 8, 9, 13, 10)) +
  labs(color = "Size", shape = "Size")
```

@@

-   All data used in this manuscript and vignettes have been cited and can be found on our repository pages.
-   Manuscript repository with code and data: <https://osf.io/swmva/> or <https://github.com/SemanticPriming/stimuli-power>
-   Package repository with vignettes and data: <https://github.com/SemanticPriming/semanticprimeR>
-   We did not pre-register this study, as it was a simulation study. No materials were used.

>>>>

In this section, we provide an example of the suggested procedure. The first dataset includes concreteness ratings from @brysbaert2014. Instructions given to participants denoted the difference between concrete (i.e., "refers to something that exists in reality") and abstract (i.e., "something you cannot experience directly through your senses or actions") terms. Participants were then asked to rate concreteness of terms using a 1 (*abstract*) to 5 (*concrete*) scale. This data represents a small scale dataset (i.e., the range of the scale of the data is small, 4 points) that could be used as pilot data for a study using concrete word ratings. The data is available at <https://osf.io/qpmf4/> [see the following for thoughts on analyzing ordinal rating data: @bürkner2019; @liddell2018; @taylor2022; @kubinec2023].

@@

For this example, imagine a researcher who wants to determine the differences in response latencies for abstract and concrete words. They will select *n* = 40 words from the rating data from @brysbaert2014 that are split evenly into abstract and concrete ends of the rating scale. In the experiment, each participant will be asked to rate the words for their concreteness, and then complete a lexical decision task with these words as the phenomenon of interest. Using both datasets and the procedure outlined above, we can determine the sample size necessary to ensure adequately measured concreteness ratings and response latencies.

@@

@@

*Step 1*. The concreteness ratings data includes `r length(unique(conDF$Word))` concepts that were rated for their concreteness. We randomly selected *n* = 20 abstract words ($M_{Rating}$ \<= 2) and *n* = 20 concrete words ($M_{Rating}$ \>= 4). In the original study, not every participant rated every word, which created uneven sample sizes for each word. Further, participants were allowed to indicate they did not know a word, and those responses were set to missing data. In our sample of 40 words, the average pilot sample size was `r round(mean(tapply(con_small$Rating, con_small$Word, length)), digits = 2)` (*SD* = `r round(sd(tapply(con_small$Rating, con_small$Word, length)), digits = 2)`), and we will use `r pilot_size_c` as our pilot sample size for the concreteness ratings (this information will be used in the follow-up to the simulation study). We first selected the same real words in the ELP data as the concreteness subset selected above, and this data includes `r length(unique(elpDF$Stimulus))` real words. The average pilot sample size for this random sample was `r format(round(mean(tapply(elp_small$RT, elp_small$Stimulus, length)), digits = 2), nsmall = 2)` (*SD* = `r round(sd(tapply(elp_small$RT, elp_small$Stimulus, length)), digits = 2)`), and *n* = `r pilot_size_e` will be our pilot size for the lexical decision task.

@@

## Special Considerations

### Bimodal Data

The above simulations assume symmetric population distributions with small, medium, or large heterogeneity or a skewed distribution with the same heterogeneity. However, as shown in @pollock2018, sample rating data that appears to have a mean distribution may represent the average two separate distributions (i.e., a bimodal distribution). To examine our procedure with bimodal distributions, we estimated thirty items for three distributions: symmetric, ceiling, and floor distributions for Likert data with small heterogeneity. Bimodal data was created by selecting half of the ceiling distribution and floor distribution to combine together. The number of bimodal items was varied from 0 percent to 100 percent increasing by 10 percent increments for simulated samples. Figure \@ref(fig:bimodal-figure) shows an example of 10 of the items within a simulation that estimated that half of the Likert items would be bimodal in nature. The researcher procedure described above was then carried out using *semanticprimeR*.

```{r import-bimodal, echo = F, warning = F, message = F}
bi_DF <- import("simulation_study/simulation_bimodal/summary_bimodal.csv") %>% 
  mutate(new_percent = rep(c(80, 85, 90, 95), times = 11)) %>% 
  mutate(size = factor(size*100))

@@

```{r bimodal-figure, fig.cap = "Simulated example of Likert data with half of the items as bimodal distributions.", fig.height=10, fig.width=7, warning = F, include = F, message = F}
ggplot(researcher_sample %>% 
         filter(item < 11), aes(x = score)) +
  geom_histogram(aes(y = ..count..), bins = 7, position = "identity", alpha = 0.7, color = "black") +  # Stacked histograms
  labs(x = "Score", y = "Item") +
  theme_minimal() +
  facet_wrap(~factor(item), scales = "free_y", nrow = 10) +  # Place each item in a separate row
  theme(legend.position = "none",
        axis.text.y = element_blank()) + 
  coord_cartesian(xlim = c(1, 7))  # Set x-axis limits
```

Figure \@ref(fig:bimodal-figure) portrays the results of the suggested procedure. The sample with 0 bimodal items demonstrates the same values as above. The proposed sample size increases from 10 to 40% because of the heterogeneity in SE across items (i.e., with 10% of items with a larger SE due to their bimodal nature, the required sample size increases). From 50% to 100% the sample size decreases because the variance of the SEs across items decreases (i.e., at 100 percent, they are all larger rather than a mix). In theory, these results map onto the conceptual framework - if all items are truly bimodal, we have precisely measured the mean of the bimodal distribution, but this result is likely not the intended result. If researchers expect these distributions (or have representative pilot data), they could examine the data for bimodal items. These items could be estimated separately (i.e., floor effects item 1, ceiling effects item 1) to ensure that both populations of answers are represented in the data.

```{r bimodal-results, fig.cap = "Estimated sample sizes for distributions that vary in the number of bimodal items. Note that percent below is treated as a continuous factor, and therefore, some estimates are the same (i.e., below 80 and 85 percent may be both estimated as 88%).", fig.height=8, fig.width=8}
ggplot(bi_DF, aes(x = percent_below,
                  y = corrected_sample_size, 
                  color = factor(size),
                  shape = factor(size))) +
  geom_point(size = 3) +
  geom_line() +
  theme_classic() +
  xlab("Percent of Items Below Threshold") + 
  ylab("Corrected Sample Size") +
  scale_color_manual(values = c("#1f77b4", "#ff7f0e", "#2ca02c", "#d62728", "#9467bd", 
                                "#8c564b", "#e377c2", "#7f7f7f", "#bcbd22", "#17becf", 
                                "#9edae5")) +  # Custom high-contrast color palette
  scale_shape_manual(values = c(16, 17, 18, 19, 15, 3, 7, 8, 9, 13, 10)) +
  labs(color = "Size", shape = "Size")
```

### Combining Tools

Researchers may often be interested in more than just the precision for individual items. The estimation of power via traditional power calculations for a statistical test or the reliability of items could also be of interest for an appropriate dataset for analyses. We would suggest that researchers combine reliability with precision to collect data that are both reliable and precisely measured. For example, one could use our proposed procedure to calculate the sample size for precision. Separately, the researcher would determine the level of reliability they would like to find in items based on previous research or practical guidelines. The minimum sample size could be set based on estimations for hypothesis testing, the stopping rule based on minimum reliability and desired SE for items, and a maximum sample size based on simulations or practical matters. Each data collection represents a unique scenario in which researchers can combine tools based on their needs to collect precise, reliable, and (traditionally defined) adequately powered data. While our proposal may bring to mind the problems with researcher degrees of freedom [@simmons2011], transparent practices decisions around sample size planning should be encouraged to limit potential questionable research practices.

@@

- Data: All data used in this manuscript and vignettes have been cited and can be found on our repository pages (https://github.com/SemanticPriming/stimuli-power).
- Analysis Code: 
  - Manuscript repository with code and data: <https://osf.io/swmva/> or <https://github.com/SemanticPriming/stimuli-power>
  - Package repository with vignettes and data: <https://github.com/SemanticPriming/semanticprimeR>
- Pre-registration: We did not pre-register this study, as it was a simulation study.
- Materials: No materials were used.
- Funding: no funding was provided for this study. 
- Conflicts of interest: the researchers declare no conflicts of interest.

### 2024-12-01 manuscript_draft.tex
updating manuscript draft

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\keywords{accuracy in parameter estimation, power, sampling, simulation, hypothesis testing}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

In this section, we provide an example of the suggested procedure. The first dataset includes concreteness ratings from Brysbaert et al. (2014). Instructions given to participants denoted the difference between concrete (i.e., ``refers to something that exists in reality'') and abstract (i.e., ``something you cannot experience directly through your senses or actions'') terms. Participants were then asked to rate concreteness of terms using a 1 (\emph{abstract}) to 5 (\emph{concrete}) scale. This data represents a small scale dataset (i.e., the range of the scale of the data is small, 4 points) that could be used as pilot data for a study using concrete word ratings. The data is available at \url{https://osf.io/qpmf4/}.

@@

For this example, imagine a researcher who wants to determine the differences in response latencies for abstract and concrete words (see the following for thoughts on analyzing this type of data: Bürkner \& Vuorre, 2019; Liddell \& Kruschke, 2018; Taylor et al., 2022). They will select \emph{n} = 40 words from the rating data from Brysbaert et al. (2014) that are split evenly into abstract and concrete ends of the rating scale. In the experiment, each participant will be asked to rate the words for their concreteness, and then complete a lexical decision task with these words as the phenomenon of interest. Using both datasets and the procedure outlined above, we can determine the sample size necessary to ensure adequately measured concreteness ratings and response latencies.

@@

\subsection{Special Consideration: Bimodal Data}\label{special-consideration-bimodal-data}

@@

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{manuscript_draft_files/figure-latex/bimodal-results-1.pdf}}
\caption{\label{fig:bimodal-results}Estimated sample sizes for distributions that vary in the number of bimodal items. Note that percent below is treated as a continuous factor}
\end{figure}

@@

\begin{itemize}
\tightlist
\item
  All data used in this manuscript and vignettes have been cited and can be found on our repository pages.
\item
  Manuscript repository with code and data: \url{https://osf.io/swmva/} or \url{https://github.com/SemanticPriming/stimuli-power}
\item
  Package repository with vignettes and data: \url{https://github.com/SemanticPriming/semanticprimeR}
\item
  We did not pre-register this study, as it was a simulation study. No materials were used.
\end{itemize}

@@

>>>>

\usepackage{xpatch}
\makeatletter
\xapptocmd\appendix
  {\xapptocmd\section
    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi: #1}}
    {}{\InnerPatchFailed}%
  }
{}{\PatchFailed}
\makeatother
\keywords{accuracy in parameter estimation, power, sampling, simulation, hypothesis testing}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

@@

In this section, we provide an example of the suggested procedure. The first dataset includes concreteness ratings from Brysbaert et al. (2014). Instructions given to participants denoted the difference between concrete (i.e., ``refers to something that exists in reality'') and abstract (i.e., ``something you cannot experience directly through your senses or actions'') terms. Participants were then asked to rate concreteness of terms using a 1 (\emph{abstract}) to 5 (\emph{concrete}) scale. This data represents a small scale dataset (i.e., the range of the scale of the data is small, 4 points) that could be used as pilot data for a study using concrete word ratings. The data is available at \url{https://osf.io/qpmf4/} (see the following for thoughts on analyzing ordinal rating data: Bürkner \& Vuorre, 2019; Kubinec, 2023; Liddell \& Kruschke, 2018; Taylor et al., 2022).

@@

For this example, imagine a researcher who wants to determine the differences in response latencies for abstract and concrete words. They will select \emph{n} = 40 words from the rating data from Brysbaert et al. (2014) that are split evenly into abstract and concrete ends of the rating scale. In the experiment, each participant will be asked to rate the words for their concreteness, and then complete a lexical decision task with these words as the phenomenon of interest. Using both datasets and the procedure outlined above, we can determine the sample size necessary to ensure adequately measured concreteness ratings and response latencies.

@@

\subsection{Special Considerations}\label{special-considerations}

\subsubsection{Bimodal Data}\label{bimodal-data}

@@

\begin{verbatim}
## Warning: The dot-dot notation (`..count..`) was deprecated in ggplot2 3.4.0.
## i Please use `after_stat(count)` instead.
## This warning is displayed once every 8 hours.
## Call `lifecycle::last_lifecycle_warnings()` to see where this warning was
## generated.
\end{verbatim}

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{manuscript_draft_files/figure-latex/bimodal-results-1.pdf}}
\caption{\label{fig:bimodal-results}Estimated sample sizes for distributions that vary in the number of bimodal items. Note that percent below is treated as a continuous factor, and therefore, some estimates are the same (i.e., below 80 and 85 percent may be both estimated as 88\%).}
\end{figure}

\subsubsection{Combine Parameters}\label{combine-parameters}

Researchers may often be interested in more than just the precision for individual items. The estimation of power via traditional power calculations for a statistical test or the reliability of items could also be of interest for an appropriate dataset for analyses. We would suggest that researchers combine reliability with precision to collect data that are both reliable and precisely measured. For example, one could use our proposed procedure to calculate the sample size for precision. Separately, the researcher would determine the level of reliability they would like to find in items based on previous research or practical guidelines. The minimum sample size could be set based on estimations for hypothesis testing, the stopping rule based on minimum reliability and desired SE for items, and a maximum sample size based on simulations or practical matters. Each data collection represents a unique scenario in which researchers can combine tools based on their needs to collect precise, reliable, and (traditionally defined) adequately powered data. While our proposal may bring to mind the problems with researcher degrees of freedom (Simmons et al., 2011), transparent practices decisions around sample size planning should be encouraged to limit potential questionable research practices.

@@

\begin{itemize}
\tightlist
\item
  Data: All data used in this manuscript and vignettes have been cited and can be found on our repository pages (\url{https://github.com/SemanticPriming/stimuli-power}).
\item
  Scripts:

  \begin{itemize}
  \tightlist
  \item
    Manuscript repository with code and data: \url{https://osf.io/swmva/} or \url{https://github.com/SemanticPriming/stimuli-power}
  \item
    Package repository with vignettes and data: \url{https://github.com/SemanticPriming/semanticprimeR}
  \end{itemize}
\item
  Pre-registration: We did not pre-register this study, as it was a simulation study.
\item
  Materials: No materials were used.
\item
  Funding: no funding was provided for this study.
\item
  Conflicts of interest: the researchers declare no conflicts of interest.
\end{itemize}

@@

\bibitem[\citeproctext]{ref-kubinec2023}
Kubinec, R. (2023). Ordered Beta Regression: A Parsimonious, Well-Fitting Model for Continuous Data with Lower and Upper Bounds. \emph{Political Analysis}, \emph{31}(4), 519--536. \url{https://doi.org/10.1017/pan.2022.20}
